Neural Sparse Gaussian Process: For Updated Exploration

### 2023-04-14 paper/main.tex
added pausl titles
added few changes data
added abstract

%\title{Investigatin Uncertainty Quantification in Model-based Reinforcement Learning}
% \title{Model-based Reinforcement Learning with Fast Posterior Updates}
\title{Sequential Decision-Making under Uncertainty with Big Data}
\author{%
  Aidan ~Scannell \\
  Aalto University\\
  %Finnish Center for Artificial Intelligence \\
  \texttt{aidan.scannell@aalto.fi}
  \And
  Paul ~Chang \\
  Aalto University\\
  \texttt{paul.chang@aalto.fi}
  \And
  Ella ~Tamir \\
  Aalto University\\
  \texttt{ella.tamir@aalto.fi}
  \And
  Arno ~Solin \\
  Aalto University\\
  \texttt{arno.solin@aalto.fi}
  \And
  Joni ~Pajarinen \\
  Aalto University\\
  \texttt{joni.pajarinen@aalto.fi}
  % examples of more authors
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

@@

\begin{abstract}
  % Reinforcement learning (RL) agents typically do not utilise the information previously gained during an episode.
  % as they perform learning offline, i.e. after an episode.
  % Reinforcement learning (RL) agents typically perform learning after an episode, ignoring the potential information gain whilst collecting data.
  % Reinforcement learning (RL) agents typically perform learning offline and do not utilise the information gained during an episode.
  % We present a Bayesian model-based RL algorithm which incorporates this information by updating a dynamic model's posterior during an episode.
  % Our method leverages the fast posterior updates available in the dual parameterisation of sparse variational Gaussian processes.
  % Importantly, our algorithm scales to 1) environment's with high dimensional state-spaces as it is built upon the connection between Bayesian neural networks (BNNs) and
  % Gaussian processes (GPs) and 2) long episodes, as the complexity of our updates is independent of episode length.
  % We demonstrate that our fast posterior updates improve sample efficiency in environments with action penalties, a notoriously difficult
  % challenge for model-based RL algorithms.
  %
  % Reinforcement learning (RL) agents typically perform learning offline and do not utilise the information gained during an episode.
  % We present a Bayesian model-based RL algorithm which incorporates this information by updating a dynamic model's posterior during an episode.
  % Typically training a neural network dynamic model is not feasible and nonparameteric methods such as Gaussian processes scale poorly with data.
  % Importantly, our algorithm scales to 1) environment's with high dimensional state-spaces, 2) large data sets and 3) long episodes,
  % as the complexity of our updates is independent of episode length.
  % We demonstrate that our fast posterior updates improve sample efficiency in environments with action penalties, a notoriously difficult
  % challenge for model-based RL algorithms.
  %
  Deep neural networks

  Obtain SVGP posterior on trained deep neural network (NN) without optimising any variational parameters.

  Dual formulation enables fast updates for 1) model-based RL 2) pathwise, function-space conditioning.
  % Reinforcement learning (RL) agents typically perform learning offline and do not utilise the information gained during an episode.
  % We present a Bayesian model-based RL algorithm which incorporates this information by updating a dynamic model's posterior during an episode.
  % Typically training a neural network dynamic model is not feasible and nonparameteric methods such as Gaussian processes scale poorly with data.
  % Importantly, our algorithm scales to 1) environment's with high dimensional state-spaces, 2) large data sets and 3) long episodes,
  % as the complexity of our updates is independent of episode length.
  % We demonstrate that our fast posterior updates improve sample efficiency in environments with action penalties, a notoriously difficult
  % challenge for model-based RL algorithms.
\end{abstract}

@@

\textbf{Deep learning}
Given the data set $\dataset = \{\mathbf{x}_{n} \in \inputDomain, \mathbf{y}_{n} \in \outputDomain\}_{n=0}^{N}$,
the goal of (supervised) deep learning, is to train the weights $\transitionParams \in \R^{P}$ of an $L\text{-layer}$ NN
$f : \inputDomain \rightarrow \outputDomain$ to minimize the (regularized) empirical risk,
\begin{align} \label{eq-empirical-risk}
  \transitionParams^{*} = \arg \min_{\transitionParams \in \R^{D}} \mathcal{L}(\dataset;\transitionParams) = \arg \min_{\transitionParams \in \R^{D}} \left(
  R(\transitionParams) + \sum_{n=0}^{N-1} l(f(\mathbf{x}_{n} ; \transitionParams), \mathbf{y}_{n}) \right).
\end{align}
In practice, it is common to use the mean squared error (MSE) loss
$l(f(\mathbf{x}_{n} ; \transitionParams),\mathbf{y}_{n}) = \|f(\mathbf{x}_{n} ; \transitionParams) - \mathbf{y}_{n} \|^{2}_{2}$
and to use the weight decay regularizer $R(\transitionParams)=\frac{1}{2}\gamma^{-2}\|\transitionParams\|^{2}_{2}$.
However, the practioner is free to choose any loss and regularizer. \todo{I think?}
% \textbf{Bayesian neural networks}
% \textbf{Weight space to function space}
From a Bayesian perspective, we can interpret the terms in \cref{eq-empirical-risk} as the log-prior and the i.i.d. log likelihood, i.e.,
\begin{align} \label{eq-log-prior}
  R(\transitionParams) &= \underbrace{\log p(\transitionParams)}_{\text{log prior}} = \log \mathcal{N}(\transitionParams \mid 0, \gamma^{2} \mathbf{I}) \\
  l(f(\mathbf{x}_{n} ; \transitionParams), \mathbf{y}_{n}) &= \underbrace{\log p(\mathbf{y}_{n} \mid f(\mathbf{x}_{n}; \transitionParams))}_{\text{log likelihood}}
  = \log \mathcal{N} \left( \mathbf{y}_{n} \mid f(\mathbf{x}_{n}; \transitionParams), \mathbf{I} \right) \label{eq-log-likelihood}
\end{align}

@@

\end{document}

>>>>

%\title{Investigatin Uncertainty Quantification in Model-based Reinforcement Learning}
% \title{Model-based Reinforcement Learning with Fast Posterior Updates}
\title{Neural Sparse Gaussian Process: For Updated Exploration}
\author{%
  Aidan ~Scannell \\
  Aalto University\\
  %Finnish Center for Artificial Intelligence \\
  \texttt{aidan.scannell@aalto.fi}
  \And
  Paul ~Chang \\
  Aalto University\\
  \texttt{paul.chang@aalto.fi}
  \And
  Ella ~Tamir \\
  Aalto University\\
  \texttt{ella.tamir@aalto.fi}
  \And
  Arno ~Solin \\
  Aalto University\\
  \texttt{arno.solin@aalto.fi}
  \And
  Joni ~Pajarinen \\
  Aalto University\\
  \texttt{joni.pajarinen@aalto.fi}
  % examples of more authors
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

@@

\begin{abstract}
  % Reinforcement learning (RL) agents typically do not utilise the information previously gained during an episode.
  % as they perform learning offline, i.e. after an episode.
  % Reinforcement learning (RL) agents typically perform learning after an episode, ignoring the potential information gain whilst collecting data.
  % Reinforcement learning (RL) agents typically perform learning offline and do not utilise the information gained during an episode.
  % We present a Bayesian model-based RL algorithm which incorporates this information by updating a dynamic model's posterior during an episode.
  % Our method leverages the fast posterior updates available in the dual parameterisation of sparse variational Gaussian processes.
  % Importantly, our algorithm scales to 1) environment's with high dimensional state-spaces as it is built upon the connection between Bayesian neural networks (BNNs) and
  % Gaussian processes (GPs) and 2) long episodes, as the complexity of our updates is independent of episode length.
  % We demonstrate that our fast posterior updates improve sample efficiency in environments with action penalties, a notoriously difficult
  % challenge for model-based RL algorithms.
  %
  % Reinforcement learning (RL) agents typically perform learning offline and do not utilise the information gained during an episode.
  % We present a Bayesian model-based RL algorithm which incorporates this information by updating a dynamic model's posterior during an episode.
  % Typically training a neural network dynamic model is not feasible and nonparameteric methods such as Gaussian processes scale poorly with data.
  % Importantly, our algorithm scales to 1) environment's with high dimensional state-spaces, 2) large data sets and 3) long episodes,
  % as the complexity of our updates is independent of episode length.
  % We demonstrate that our fast posterior updates improve sample efficiency in environments with action penalties, a notoriously difficult
  % challenge for model-based RL algorithms.
  %
  %Models that can process new information quickly

 % Fast adaptation of 
 Building systems that adapt to new information is essential to deploy methods in real-world applications. While gradient-based training is symbiotic to modern deep learning, it is rigid regarding adding further information, usually involving retraining from scratch. In a functional view, we can view data as the parameters of our model instead of the weights, which we call the dual parameterization of a model. Using the formulation we can assimilate data without gradient-based retraining in dual parameterization. Our proposed method allows for deployment in any sequential decision-making process. We show its effectiveness in model-based reinforcement learning using updated information for more efficient exploration.

  %Deep neural networks

%  Obtain SVGP posterior on trained deep neural network (NN) without optimising any variational parameters.

  %Dual formulation enables fast updates for 1) model-based RL 2) pathwise, function-space conditioning.
  % Reinforcement learning (RL) agents typically perform learning offline and do not utilise the information gained during an episode.
  % We present a Bayesian model-based RL algorithm which incorporates this information by updating a dynamic model's posterior during an episode.
  % Typically training a neural network dynamic model is not feasible and nonparameteric methods such as Gaussian processes scale poorly with data.
  % Importantly, our algorithm scales to 1) environment's with high dimensional state-spaces, 2) large data sets and 3) long episodes,
  % as the complexity of our updates is independent of episode length.
  % We demonstrate that our fast posterior updates improve sample efficiency in environments with action penalties, a notoriously difficult
  % challenge for model-based RL algorithms.
\end{abstract}

@@

\textbf{Deep learning}
Given the data set $\dataset = \{\mathbf{x}_{n} \in \inputDomain, \mathbf{y}_{n} \in \outputDomain\}_{n=0}^{N}$,
the goal of (supervised) deep learning, is to train the weights $\transitionParams \in \R^{P}$ of an $L\text{-layer}$ NN
$f : \inputDomain \rightarrow \outputDomain$ to minimize the (regularized) empirical risk,
\begin{align} \label{eq-empirical-risk}
  \transitionParams^{*} = \arg \min_{\transitionParams \in \R^{D}} \mathcal{L}(\dataset;\transitionParams) = \arg \min_{\transitionParams \in \R^{D}} \left(
  \sum_{n=0}^{N-1} l(f(\mathbf{x}_{n} ; \transitionParams), \mathbf{y}_{n}) + R(\transitionParams)  \right).
\end{align}
In practice, it is common to use the mean squared error (MSE) loss
$l(f(\mathbf{x}_{n} ; \transitionParams),\mathbf{y}_{n}) = \|f(\mathbf{x}_{n} ; \transitionParams) - \mathbf{y}_{n} \|^{2}_{2}$
and to use the weight decay regularizer $R(\transitionParams)=\frac{1}{2}\gamma^{-2}\|\transitionParams\|^{2}_{2}$.
However, the practioner is free to choose any loss and regularizer. \todo{I think?}
% \textbf{Bayesian neural networks}
% \textbf{Weight space to function space}
From a Bayesian perspective, we can interpret the terms in \cref{eq-empirical-risk} as the log-prior and the i.i.d. log likelihood, i.e.,
\begin{align} \label{eq-log-prior}
  R(\transitionParams) &= \underbrace{\log p(\transitionParams)}_{\text{log prior}} = \log \mathcal{N}(\transitionParams \mid 0, \gamma^{2} \mathbf{I}) \\
  l(f(\mathbf{x}_{n} ; \transitionParams), \mathbf{y}_{n}) &= \underbrace{\log p(\mathbf{y}_{n} \mid f(\mathbf{x}_{n}; \transitionParams))}_{\text{log likelihood}}
  = \log \mathcal{N} \left( \mathbf{y}_{n} \mid f(\mathbf{x}_{n}; \transitionParams), \mathbf{I} \right) \label{eq-log-likelihood}
\end{align}

@@

%\bibliography{biblio.bib}
\end{document}
